---
title: "tract block"
author: "Lilac Hong"
date: "2026-04-30"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: true
---

```{r include = FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)

here::i_am("code/2_2_tract_ndvi.Rmd")
```

# Tract-Level Analysis: Housing Growth, Demographic Change, and Greenness Exposure, 2010–2020

Companion output to `code/2_2_tract_ndvi.R`. Tract-level analog of `1_2_block_ndvi.R`, using the minimum-viable scenario set recommended by `code/2_1_tract_ndvi_scenario_comparison.Rmd`.

## Guiding questions

1. **Where is housing being built and who lives there** (and how have demographics changed over time)?
2. **Who is experiencing changes in exposure to green environments?**

---

## What changed vs. the block-level analysis

| Dimension | Block (`1_2_block_ndvi.R`) | Tract (`2_2_tract_ndvi.R`) |
|---|---|---|
| Spatial unit | Census block (~83k rows per scenario) | Census tract (~1,318 rows per scenario) |
| Response variables | NDVI only | NDVI only |
| Seasons retained | Dormant, Growing | Dormant, Growing |
| Buffers retained | Dormant: 1000 m; Growing: 500/1000/2000 m | **1000 m only** for both seasons |
| Total scenarios | 4 | **2** |
| Baseline R² range | 0.006–0.018 | **0.425–0.509** |

### Why only 1000 m at the tract level?

From `code/2_1_tract_ndvi_scenario_comparison.md`:

1. **Buffer main effect is not statistically significant at the tract level.** Nested F-tests for buffer (per season) yield p = 0.10 (dormant) and p = 0.39 (growing).
2. **Cross-buffer ΔGHI correlations are very high.** Dormant: r = 0.96–0.99. Growing: r = 0.75–0.99.
3. **Cohen's d for 500 vs. 2000 m is negligible** (|d| ≤ 0.07).
4. **AIC/BIC prefer `* season` over `+ season + buffer`.** Adding buffer actively hurts BIC at the tract scale.

---

```{r echo=FALSE}
# -----------------------------
# 1. READ DATA
# -----------------------------
ndvi_tract <- read.csv(here::here("data", "tract_final.csv"))

# -----------------------------
# 2. BASIC CLEANING
# -----------------------------
df <- ndvi_tract %>%
  mutate(
    across(all_of(c("NDVI", "HU_density")), as.numeric),
    year   = as.integer(year),
    buffer = as.integer(buffer)
  ) %>%
  filter(year %in% c(2010, 2020)) %>%
  filter(!is.na(NDVI), !is.na(HU_density), HU_density > 0) %>%
  mutate(log_HU_density = log(HU_density))

# -----------------------------
# 3. PIVOT TO WIDE
# -----------------------------
wide <- df %>%
  select(tract, season, buffer, year, log_HU_density, NDVI, HU, POP, area_ha) %>%
  pivot_wider(names_from  = year,
              values_from = c(log_HU_density, NDVI, HU, POP),
              names_sep   = "_") %>%
  drop_na() %>%
  mutate(delta_log_HU = log_HU_density_2020 - log_HU_density_2010,
         delta_NDVI   = NDVI_2020 - NDVI_2010,
         delta_HU     = HU_2020 - HU_2010,
         delta_POP    = POP_2020 - POP_2010,
         HU_density_2010 = HU_2010 / area_ha,
         HU_density_2020 = HU_2020 / area_ha,
         delta_HU_density = HU_density_2020 - HU_density_2010)

# -----------------------------
# 4. RESTRICT TO UPDATED SCENARIO SET
# -----------------------------
scenarios <- tibble::tribble(
  ~season,           ~buffer,
  "Dormant_Season",     1000,
  "Growing_Season",     1000
)

wide <- wide %>% semi_join(scenarios, by = c("season", "buffer"))

# ============================================================
# 8. VISUALIZATIONS (GREENNESS)
# ============================================================
theme_lambert <- theme_minimal(base_size = 11) +
  theme(strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        legend.position  = "bottom")

```

## Part I: Pseudo-Residual Greenness Analysis

### Methods

Same as the block analysis, applied to `data/tract_final.csv`:

$$\text{NDVI}_{2010} = \alpha + \beta \cdot \log(\text{HU_density}_{2010})$$

- GHI~2010~ = NDVI~2010~ − ($\hat{\alpha}_{2010}$ +$\hat{\beta}_{2010}$ · log($\text{HU_density}_{2010}$))
- GHI~2020~= NDVI~2020~ − ($\hat{\alpha}_{2010}$ +$\hat{\beta}_{2010}$ · log($\text{HU_density}_{2020}$)))
- **ΔGHI = GHI~2020~ − GHI~2010~**

Sample after filtering (NDVI present, HU_density > 0, both years present): **2,636 tract × scenario rows** across **1,318 unique Washington census tracts**.

### Table 1 — 2010 baseline regressions

```{r echo=FALSE}
# ============================================================
# 5. FIT 2010 BASELINE REGRESSIONS
# ============================================================
results <- scenarios %>%
  mutate(data = map2(season, buffer, ~ wide %>%
                       filter(season == .x, buffer == .y))) %>%
  mutate(mod_green = map(data, ~ lm(NDVI_2010 ~ log_HU_density_2010, data = .x)))

reg_table <- results %>%
  mutate(
    tidy_green = map(mod_green, ~ {
      s <- summary(.x)
      tibble(
        intercept = coef(.x)[1],
        slope     = coef(.x)[2],
        slope_se  = coef(s)[2, 2],
        slope_p   = coef(s)[2, 4],
        R2        = s$r.squared,
        n         = nobs(.x)
      )
    })
  ) %>%
  select(season, buffer, tidy_green) %>%
  unnest(tidy_green)

cat("TABLE 1: 2010 OLS Regression Coefficients (Tract-Level Greenness)\n")
cat("  NDVI_2010 ~ log(HU_density_2010)\n")
print(as.data.frame(reg_table), digits = 4, row.names = FALSE)
```

| Scenario | n | Intercept | Slope | Slope SE | p | R² |
|---|---:|---:|---:|---:|---:|---:|
| Dormant × 1000 m | 1,318 | 0.5458 | −0.04358 | 0.00140 | <1e-16 | 0.4248 |
| Growing × 1000 m | 1,318 | 0.6384 | −0.04908 | 0.00133 | <1e-16 | 0.5092 |

**Read:** At the tract scale the housing-density → NDVI relationship is strongly negative (~−0.045 NDVI units per log unit of HU density/ha), and baseline $R^2$ is ~0.43–0.51. This is a fundamentally different picture than at the block level (slopes near zero, $R^2$ < 0.02). Aggregation reveals what block-level noise obscures: denser tracts are measurably less green.

### Table 2 — ΔGHI summary

```{r echo=FALSE}
# ============================================================
# 6. COMPUTE RESIDUALS AND PSEUDO-RESIDUALS
# ============================================================
results <- results %>%
  mutate(
    data = map2(data, mod_green, function(d, mg) {
      d %>%
        mutate(
          resid_green_2010 = NDVI_2010 - predict(mg, newdata = d),
          pred_green_2020  = predict(mg, newdata = data.frame(
            log_HU_density_2010 = d$log_HU_density_2020)),
          resid_green_2020 = NDVI_2020 - pred_green_2020,
          change_green     = resid_green_2020 - resid_green_2010
        )
    })
  )

all_data <- results %>%
  select(data) %>%
  unnest(data) %>%
  mutate(
    state_green_2010 = ifelse(resid_green_2010 >= 0, "Bright", "Dim"),
    state_green_2020 = ifelse(resid_green_2020 >= 0, "Bright", "Dim"),
    transition_green = paste0(state_green_2010, " -> ", state_green_2020),
    direction_green  = case_when(
      change_green > 0 ~ "Brightened",
      change_green < 0 ~ "Dimmed",
      TRUE ~ "No change"
    ),
    scenario = factor(paste(ifelse(season == "Dormant_Season",
                                   "Dormant", "Growing"),
                            "1000 m"),
                      levels = c("Dormant 1000 m", "Growing 1000 m"))
  )


# ============================================================
# TABLE 2: Change summary by scenario
# ============================================================
change_summary <- all_data %>%
  group_by(season, buffer) %>%
  summarise(
    n                    = n(),
    mean_change_green    = mean(change_green),
    sd_change_green      = sd(change_green),
    median_change_green  = median(change_green),
    pct_brightened_green = mean(change_green > 0) * 100,
    pct_dimmed_green     = mean(change_green < 0) * 100,
    .groups = "drop"
  )

cat("TABLE 2: Tract-Level Greenness Change Summary by Scenario\n")
print(as.data.frame(change_summary), digits = 4, row.names = FALSE)
```


| Scenario | n | Mean | SD | Median | % Brightened | % Dimmed |
|---|---:|---:|---:|---:|---:|---:|
| Dormant × 1000 m | 1,318 | **−0.0862** | 0.0992 | −0.0694 | 19.9 | 80.1 |
| Growing × 1000 m | 1,318 | **+0.0106** | 0.0189 | +0.0114 | 77.8 | 22.2 |

**Read:**
- Dormant: strong net dimming (mean ≈ −0.086 NDVI units; only ~20% of tracts brighten). Magnitude is nearly identical to block-level (block mean ≈ −0.083).
- Growing: mild net brightening (mean ≈ +0.011; ~78% of tracts brighten). Magnitude is roughly double the block-level growing signal (+0.006), suggesting tract aggregation amplifies the growing-season brightening signal.

### Table 3 — Bright/Dim transitions, 2010 → 2020

```{r echo=FALSE}
# ============================================================
# TABLE 3: Bright/Dim transitions
# ============================================================
transition_green <- all_data %>%
  group_by(season, buffer, transition_green) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = transition_green, values_from = n, values_fill = 0)

cat("TABLE 3: Greenness Bright/Dim Transitions (2010 -> 2020)\n")
print(as.data.frame(transition_green), row.names = FALSE)
```


**Read:**
- Dormant: Bright → Dim (495) outnumbers Dim → Bright (57) by ~8.7× — even sharper asymmetry than at the block level (~7.5×).
- Growing: Bright → Bright (759) dominates. Most tracts that were already greener-than-expected in the growing season held that status.

### Paired t-test — Dormant × 1000 m vs. Growing × 1000 m

```{r echo=FALSE}
# ============================================================
# 7. STATISTICAL COMPARISONS
# ============================================================

# Paired t-test: Dormant vs Growing (same tract, 1000 m)
cat("PAIRED t-TEST: Dormant_Season_1000 vs Growing_Season_1000\n")

paired <- all_data %>%
  select(tract, season, change_green) %>%
  pivot_wider(names_from = season, values_from = change_green) %>%
  drop_na()

# print(t.test(paired$Dormant_Season, paired$Growing_Season, paired = TRUE))
```

- n = 1,318 tracts
- Mean difference (Dormant − Growing) = **−0.0968**
- t = −34.87, p < 2.2e-16

The same tract's dormant vs. growing ΔGHI differ by ~0.10 units on average. Dormant season is dimming; growing season is brightening.

---

## Part II: Where Is Housing Being Built and Who Lives There?

This section directly addresses the first guiding question. The previous version of this analysis used housing density only as a regression predictor. Here we characterize housing growth itself — its magnitude, spatial distribution, and demographic context.

```{r echo=FALSE}
# ============================================================
# 9. DEMOGRAPHIC CHANGE: RACE & AGE
# ============================================================

age  <- read.csv(here::here("data", "age", "age_10_20.csv"))
race <- read.csv(here::here("data", "race", "race_10_20.csv"))

# Normalize tract ID for joins (strip anything non-numeric)
age$tract  <- as.numeric(gsub("[^0-9]", "", as.character(age$tract20l)))
race$tract <- as.numeric(gsub("[^0-9]", "", as.character(race$tract20l)))

age_bands <- c("age0_4","age5_9","age10_14","age15_19","age20_24","age25_29",
               "age30_34","age35_39","age40_44","age45_49","age50_54","age55_59",
               "age60_64","age65_69","age70_74","age75_79","age80_84","age85_up")
band_mid  <- c(2.5, 7.5, 12.5, 17.5, 22.5, 27.5, 32.5, 37.5, 42.5,
               47.5, 52.5, 57.5, 62.5, 67.5, 72.5, 77.5, 82.5, 90.0)

# --- Age summary variables ---
age_summary <- function(df, yr) {
  cols <- paste0(age_bands, "_", yr)
  A <- as.matrix(df[, cols])
  total <- rowSums(A)
  total_safe <- ifelse(total == 0, NA_real_, total)
  P <- A / total_safe
  mean_age     <- P %*% band_mid
  pct_under15  <- rowSums(A[, 1:3]) / total_safe * 100
  pct_65p      <- rowSums(A[, 14:18]) / total_safe * 100
  data.frame(total = total,
             mean_age = as.numeric(mean_age),
             pct_under15 = pct_under15,
             pct_65p = pct_65p)
}
a10 <- age_summary(age, 2010)
a20 <- age_summary(age, 2020)

# --- Age delta-chi ---
age_delta_chi <- function(df) {
  A10 <- as.matrix(df[, paste0(age_bands, "_2010")])
  A20 <- as.matrix(df[, paste0(age_bands, "_2020")])
  P10 <- A10 / ifelse(rowSums(A10) == 0, NA_real_, rowSums(A10))
  P20 <- A20 / ifelse(rowSums(A20) == 0, NA_real_, rowSums(A20))
  Pbar <- (P10 + P20) / 2
  num  <- (P20 - P10)^2
  ratio <- ifelse(Pbar > 0, num / Pbar, 0)
  rowSums(ratio, na.rm = TRUE)
}

age_out <- data.frame(
  tract              = age$tract,
  total_pop_2010     = a10$total,
  total_pop_2020     = a20$total,
  mean_age_2010      = a10$mean_age,
  mean_age_2020      = a20$mean_age,
  pct_under15_2010   = a10$pct_under15,
  pct_under15_2020   = a20$pct_under15,
  pct_65p_2010       = a10$pct_65p,
  pct_65p_2020       = a20$pct_65p,
  delta_chi_age      = age_delta_chi(age)
)
age_out$delta_mean_age    <- age_out$mean_age_2020    - age_out$mean_age_2010
age_out$delta_pct_under15 <- age_out$pct_under15_2020 - age_out$pct_under15_2010
age_out$delta_pct_65p     <- age_out$pct_65p_2020     - age_out$pct_65p_2010
age_out$delta_pop         <- age_out$total_pop_2020   - age_out$total_pop_2010

# --- Race summary + delta-chi ---
race_cats <- c("AIAN", "ASIAN", "Black", "NHOPI", "Two", "White")
R10 <- as.matrix(race[, paste0(race_cats, "2010")])
R20 <- as.matrix(race[, paste0(race_cats, "2020")])
T10 <- rowSums(R10); T20 <- rowSums(R20)
P10 <- R10 / ifelse(T10 == 0, NA_real_, T10)
P20 <- R20 / ifelse(T20 == 0, NA_real_, T20)
Pbar_race <- (P10 + P20) / 2
num_race  <- (P20 - P10)^2
ratio_race <- ifelse(Pbar_race > 0, num_race / Pbar_race, 0)

race_out <- data.frame(
  tract             = race$tract,
  total_race_2010   = T10,
  total_race_2020   = T20,
  pct_white_2010    = P10[, "White2010"] * 100,
  pct_white_2020    = P20[, "White2020"] * 100,
  delta_chi_race    = rowSums(ratio_race, na.rm = TRUE)
)
race_out$pct_non_white_2010 <- 100 - race_out$pct_white_2010
race_out$pct_non_white_2020 <- 100 - race_out$pct_white_2020
race_out$delta_pct_white     <- race_out$pct_white_2020     - race_out$pct_white_2010
race_out$delta_pct_non_white <- race_out$pct_non_white_2020 - race_out$pct_non_white_2010

demo <- age_out %>%
  full_join(race_out, by = "tract")

# Merge demographics onto the analytical dataset (ΔGHI per scenario)
analysis_demo <- all_data %>%
  left_join(demo, by = "tract")

# write.csv(demo,
#           file.path(out_dir, "tract_demographics_with_delta_chi.csv"),
#           row.names = FALSE)

# ============================================================
# 10. WHERE IS HOUSING BEING BUILT?
#
# Guiding question 1: "Where is housing being built and who
# lives there (and how have demographics changed over time)?"
#
# We characterize housing growth at the tract level, then
# cross-tabulate with demographics.
# ============================================================
```

### Table 4 — Housing growth summary (all tracts in the analytical sample)

```{r echo=FALSE}
# --- TABLE 4: Housing growth summary (all tracts) ---
# Use one row per tract (take from the dormant scenario since
# HU/POP are the same across seasons)
tract_housing <- analysis_demo %>%
  filter(season == "Dormant_Season") %>%
  select(tract, area_ha, HU_2010, HU_2020, POP_2010, POP_2020,
         delta_HU, delta_POP, delta_HU_density,
         HU_density_2010, HU_density_2020,
         starts_with("pct_white"), starts_with("pct_non_white"),
         starts_with("mean_age"), starts_with("pct_under15"),
         starts_with("pct_65p"), starts_with("delta_"),
         starts_with("total_pop"))

housing_summary <- tract_housing %>%
  summarise(
    n_tracts             = n(),
    total_HU_2010        = sum(HU_2010, na.rm = TRUE),
    total_HU_2020        = sum(HU_2020, na.rm = TRUE),
    total_delta_HU       = sum(delta_HU, na.rm = TRUE),
    pct_tracts_gained_HU = mean(delta_HU > 0, na.rm = TRUE) * 100,
    pct_tracts_lost_HU   = mean(delta_HU < 0, na.rm = TRUE) * 100,
    median_delta_HU      = median(delta_HU, na.rm = TRUE),
    mean_delta_HU        = mean(delta_HU, na.rm = TRUE),
    q25_delta_HU         = quantile(delta_HU, 0.25, na.rm = TRUE),
    q75_delta_HU         = quantile(delta_HU, 0.75, na.rm = TRUE),
    total_POP_2010       = sum(POP_2010, na.rm = TRUE),
    total_POP_2020       = sum(POP_2020, na.rm = TRUE),
    total_delta_POP      = sum(delta_POP, na.rm = TRUE)
  )

# cat("TABLE 4: Housing Growth Summary (All Tracts)\n")
# print(as.data.frame(housing_summary), digits = 4, row.names = FALSE)

```

Computed from the 1,318 tracts with valid NDVI and HU_density in both years.

| Variable | Value |
|---|---:|
| n tracts | 1,318 |
| Total HU, 2010 | 2196125 |
| Total HU, 2020 | 2451417 |
| Total ΔHU | 255292 |
| % tracts that gained HU | 91.81 |
| % tracts that lost HU | 8.194 |
| Median ΔHU per tract | 101.7 |
| Mean ΔHU per tract | 193.7 |
| Q25 ΔHU | 36.07 |
| Q75 ΔHU | 248.3 |
| Total POP, 2010 | 5102531 |
| Total POP, 2020 | 5902734 |
| Total ΔPOP | 800203 |


**Read:** The table structure captures the statewide housing growth picture. This table provides the baseline context: how much housing was added across Washington tracts between 2010 and 2020, and whether growth was concentrated or diffuse.

### Table 5 — Demographics by housing growth quartile

```{r echo=FALSE}
# --- Classify tracts by housing growth intensity ---
# Quartiles of ΔHU
tract_housing <- tract_housing %>%
  mutate(
    hu_growth_q = cut(delta_HU,
                      breaks = quantile(delta_HU,
                                        probs = c(0, 0.25, 0.5, 0.75, 1),
                                        na.rm = TRUE),
                      labels = c("Q1 (least growth)",
                                 "Q2",
                                 "Q3",
                                 "Q4 (most growth)"),
                      include.lowest = TRUE)
  )

# --- TABLE 5: Demographics by housing growth quartile ---
# "Who lives in the tracts where housing is being built?"
demo_by_hu_growth <- tract_housing %>%
  group_by(hu_growth_q) %>%
  summarise(
    n_tracts              = n(),
    mean_delta_HU         = mean(delta_HU, na.rm = TRUE),
    median_delta_HU       = median(delta_HU, na.rm = TRUE),
    # 2020 demographics (who lives there NOW)
    mean_pct_white_2020   = mean(pct_white_2020, na.rm = TRUE),
    mean_pct_nonwhite_2020 = mean(pct_non_white_2020, na.rm = TRUE),
    mean_mean_age_2020    = mean(mean_age_2020, na.rm = TRUE),
    mean_pct_under15_2020 = mean(pct_under15_2020, na.rm = TRUE),
    mean_pct_65p_2020     = mean(pct_65p_2020, na.rm = TRUE),
    # Demographic CHANGE
    mean_delta_pct_white  = mean(delta_pct_white, na.rm = TRUE),
    mean_delta_pct_nonwhite = mean(delta_pct_non_white, na.rm = TRUE),
    mean_delta_mean_age   = mean(delta_mean_age, na.rm = TRUE),
    mean_delta_pct_65p    = mean(delta_pct_65p, na.rm = TRUE),
    mean_delta_pop        = mean(delta_pop, na.rm = TRUE),
    # 2010 demographics (who lived there BEFORE)
    mean_pct_white_2010   = mean(pct_white_2010, na.rm = TRUE),
    mean_pct_nonwhite_2010 = mean(pct_non_white_2010, na.rm = TRUE),
    mean_mean_age_2010    = mean(mean_age_2010, na.rm = TRUE),
    .groups = "drop"
  )

# cat("TABLE 5: Demographics by Housing Growth Quartile\n")
# cat("  (Who lives where housing is being built?)\n")
# print(as.data.frame(demo_by_hu_growth), digits = 3, row.names = FALSE)
```

Tracts are divided into quartiles by ΔHU (housing units gained, 2010–2020). For each quartile, we report who lives there in 2020, who lived there in 2010, and how demographics changed.

| Variable | Q1 (least growth) | Q2 | Q3 | Q4 (most growth) |
|---|---:|---:|---:|---:|
| n tracts | 330 | 329 | 329 | 330 |
| Mean ΔHU | -0.305 | 66.254 | 160.646 | 547.705 |
| **2020 demographics** | | | | |
| Mean % non-White (2020) | 27.7 | 26.3 | 28.1 | 34.3 |
| Mean age (2020) | 40.4 | 40.9 | 39.6 | 37.4 |
| Mean % under 15 (2020) | 17.7 | 17.4 | 17.4 | 16.2 |
| Mean % aged 65+ (2020) | 17.4 | 18.3 | 16.4 | 13.7 |
| **Demographic change** | | | | |
| Mean Δ % non-White | 8.53 | 7.82 | 8.74 | 11.00 |
| Mean Δ mean age | 1.785 | 1.950 | 1.961 | 0.867 |
| Mean Δ % aged 65+ | 4.26 | 4.91 | 4.39 | 2.73 |
| Mean ΔPOP | 190 | 326 | 555 | 1356 |
| **2010 demographics** | | | | |
| Mean % non-White (2010) | 19.2 | 18.5 | 19.4 | 23.3 |
| Mean age (2010) | 38.6 | 38.9 | 37.6 | 36.5 |

**Read:** This table answers "who lives in the tracts where housing is being built?" If Q4 (most growth) tracts are disproportionately non-White or younger, that tells us housing construction is concentrated in communities of color or younger neighborhoods. Comparing 2010 vs. 2020 demographics within each quartile reveals whether the demographic composition of high-growth areas shifted over the decade.

```{r echo=FALSE}
# FIGURE 11: Demographics of housing growth quartiles
# Bar chart showing who lives in high-growth vs low-growth tracts
demo_hu_long <- demo_by_hu_growth %>%
  select(hu_growth_q,
         mean_pct_nonwhite_2010, mean_pct_nonwhite_2020,
         mean_mean_age_2010, mean_mean_age_2020) %>%
  pivot_longer(-hu_growth_q, names_to = "variable", values_to = "value") %>%
  mutate(
    year = ifelse(grepl("2010", variable), "2010", "2020"),
    measure = case_when(
      grepl("nonwhite", variable) ~ "% Non-White",
      grepl("age", variable)      ~ "Mean Age (years)"
    )
  )

fig11 <- ggplot(demo_hu_long,
                aes(x = hu_growth_q, y = value, fill = year)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = c("2010" = "#bdbdbd", "2020" = "#636363"),
                    name = "Year") +
  facet_wrap(~ measure, scales = "free_y") +
  labs(title    = "Demographics of Housing Growth Quartiles",
       subtitle = "Who lives in tracts that gained the most housing?",
       x = "Housing growth quartile (ΔHU, 2010-2020)",
       y = NULL) +
  theme_lambert +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

fig11
```

### Table 6 — Correlations: housing growth × demographics

```{r echo=FALSE}
# --- TABLE 6: Correlations between housing growth and demographics ---
hu_demo_cors <- tract_housing %>%
  summarise(
    n = n(),
    cor_dHU_pctWhite2010     = cor(delta_HU, pct_white_2010, use = "complete.obs"),
    cor_dHU_pctNonWhite2010  = cor(delta_HU, pct_non_white_2010, use = "complete.obs"),
    cor_dHU_meanAge2010      = cor(delta_HU, mean_age_2010, use = "complete.obs"),
    cor_dHU_pct65p2010       = cor(delta_HU, pct_65p_2010, use = "complete.obs"),
    cor_dHU_pctWhite2020     = cor(delta_HU, pct_white_2020, use = "complete.obs"),
    cor_dHU_pctNonWhite2020  = cor(delta_HU, pct_non_white_2020, use = "complete.obs"),
    cor_dHU_meanAge2020      = cor(delta_HU, mean_age_2020, use = "complete.obs"),
    cor_dHU_pct65p2020       = cor(delta_HU, pct_65p_2020, use = "complete.obs"),
    cor_dHU_dpctWhite        = cor(delta_HU, delta_pct_white, use = "complete.obs"),
    cor_dHU_dpctNonWhite     = cor(delta_HU, delta_pct_non_white, use = "complete.obs"),
    cor_dHU_dMeanAge         = cor(delta_HU, delta_mean_age, use = "complete.obs"),
    cor_dHU_dpct65p          = cor(delta_HU, delta_pct_65p, use = "complete.obs"),
    cor_dHU_dchiAge          = cor(delta_HU, delta_chi_age, use = "complete.obs"),
    cor_dHU_dchiRace         = cor(delta_HU, delta_chi_race, use = "complete.obs")
  )

# cat("TABLE 6: Correlations — Housing Growth × Demographics\n")
# print(as.data.frame(t(hu_demo_cors)), digits = 3)
```

Pearson correlations between ΔHU and demographic variables, computed across all 1,318 tracts.

| Variable | r(ΔHU, variable) |
|---|---:|
| % White (2010) | -0.1216 |
| % non-White (2010) | 0.1216 |
| Mean age (2010) | -0.1550 |
| % aged 65+ (2010) | -0.1278 |
| % White (2020) | -0.1930 |
| % non-White (2020) | 0.1930 |
| Mean age (2020) | -0.2686 |
| % aged 65+ (2020) | -0.2290 |
| Δ % White | -0.2418 |
| Δ % non-White | 0.2418 |
| Δ mean age | -0.3446 |
| Δ % aged 65+ | -0.2958 |
| Δ$\chi^2$ (Age) | 0.0577 |
| Δ$\chi^2$ (Race) | 0.3040 |

**Read:** These correlations reveal whether housing growth is systematically associated with particular demographic profiles. A positive correlation between ΔHU and % non-White would indicate that housing is being built disproportionately in communities of color. A positive correlation with Δχ² would indicate that high-growth tracts also experienced the most demographic reshuffling.

---

## Part III: Who Is Experiencing Changes in Exposure to Green Environments?

This section directly addresses the second guiding question. The previous version reported only correlations between ΔGHI and demographic change variables (the old Table 5). That approach was limited: it measured whether tracts that *changed* demographically also changed in greenness, but it did not answer who is *exposed* to the greenness change. A tract can dim without changing its demographics — the same people experience the loss.

The new analysis stratifies greenness outcomes by the demographic composition of the tract, answering: are non-White tracts, or older tracts, disproportionately experiencing dimming?

```{r echo=FALSE}
# ============================================================
# 11. WHO IS EXPERIENCING CHANGES IN GREEN EXPOSURE?
#
# Guiding question 2: "Who is experiencing changes in exposure
# to green environments?"
#
# Stratify ΔGHI by demographic characteristics of the tract.
# Use 2020 demographics (who lives there when the change is
# realized) and 2010 demographics (who was there at baseline).
# ============================================================

```

### Table 7 — ΔGHI by race composition tercile (2020 % non-White)

```{r echo=FALSE}
# --- TABLE 7: ΔGHI by race composition tercile (2020) ---
# Classify tracts by 2020 % non-White into terciles
analysis_demo <- analysis_demo %>%
  mutate(
    race_tercile_2020 = cut(pct_non_white_2020,
                            breaks = quantile(pct_non_white_2020,
                                              probs = c(0, 1/3, 2/3, 1),
                                              na.rm = TRUE),
                            labels = c("Low non-White",
                                       "Mid non-White",
                                       "High non-White"),
                            include.lowest = TRUE),
    age_tercile_2020 = cut(mean_age_2020,
                           breaks = quantile(mean_age_2020,
                                             probs = c(0, 1/3, 2/3, 1),
                                             na.rm = TRUE),
                           labels = c("Youngest",
                                      "Middle-aged",
                                      "Oldest"),
                           include.lowest = TRUE)
  )

dghi_by_race <- analysis_demo %>%
  filter(!is.na(race_tercile_2020)) %>%
  group_by(scenario, race_tercile_2020) %>%
  summarise(
    n                = n(),
    mean_DGHI        = mean(change_green, na.rm = TRUE),
    sd_DGHI          = sd(change_green, na.rm = TRUE),
    median_DGHI      = median(change_green, na.rm = TRUE),
    pct_brightened    = mean(change_green > 0, na.rm = TRUE) * 100,
    mean_pct_nonwhite = mean(pct_non_white_2020, na.rm = TRUE),
    mean_NDVI_2010   = mean(NDVI_2010, na.rm = TRUE),
    mean_NDVI_2020   = mean(NDVI_2020, na.rm = TRUE),
    mean_delta_NDVI  = mean(delta_NDVI, na.rm = TRUE),
    .groups = "drop"
  )

# cat("TABLE 7: ΔGHI by Race Composition Tercile (2020)\n")
# cat("  (Who is experiencing greenness change?)\n")
# print(as.data.frame(dghi_by_race), digits = 3, row.names = FALSE)

```

Tracts are divided into terciles by their 2020 % non-White population. For each tercile and scenario, we report the greenness change distribution.

| Scenario | Race tercile | n | Mean ΔGHI | SD | Median | % Brightened | Mean % non-White | Mean NDVI 2010 | Mean NDVI 2020 | Mean ΔNDVI |
|---|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| Dormant 1000 m | Low non-White | 440 | -0.08024 | 0.1029 | -0.0588 | 24.8 | 13.7 | 0.569 | 0.485 | -0.08372 |
| Dormant 1000 m | Mid non-White | 439 | -0.09263 | 0.1008 | -0.0785 | 18.7 | 26.1 | 0.483 | 0.385 | -0.09741 |
| Dormant 1000 m | High non-White | 439 | -0.08565 | 0.0933 | -0.0689 | 16.2 | 47.6 | 0.450 | 0.358 | -0.09231 |
| Growing 1000 m | Low non-White | 440 | 0.00938 | 0.0194 | 0.0113 | 80.5 | 13.7 | 0.663 | 0.669 | 0.00547 |
| Growing 1000 m | Mid non-White | 439 | 0.01165 | 0.0168 | 0.0120 | 76.5 | 26.1 | 0.565 | 0.571 | 0.00627 |
| Growing 1000 m | High non-White | 439 | 0.01085 | 0.0204 | 0.0102 | 76.3 | 47.6 | 0.534 | 0.538 | 0.00335 |

**Read:** If high non-White tracts show more negative ΔGHI (more dimming) or lower % brightened, that indicates a disproportionate greenness burden on communities of color. The table also reports baseline NDVI levels to assess whether these tracts started with less greenness — a "double disadvantage" if they also dim more.

```{r echo=FALSE}
# FIGURE 7: ΔGHI by race tercile (boxplot)
fig7 <- analysis_demo %>%
  filter(!is.na(race_tercile_2020)) %>%
  ggplot(aes(x = race_tercile_2020, y = change_green,
             fill = race_tercile_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = c("Low non-White"  = "#fee8c8",
                                "Mid non-White"  = "#fdbb84",
                                "High non-White" = "#e34a33"),
                    guide = "none") +
  facet_wrap(~ scenario) +
  labs(title    = "Greenness Change by Tract Racial Composition (2020)",
       subtitle = "Who is experiencing dimming vs. brightening?",
       x = "Tract race composition tercile (2020 % non-White)",
       y = "ΔGHI (pseudo-residual change)") +
  theme_lambert

fig7
```

### Table 8 — ΔGHI by age composition tercile (2020 mean age)

```{r echo=FALSE}
# --- TABLE 8: ΔGHI by age composition tercile (2020) ---
dghi_by_age <- analysis_demo %>%
  filter(!is.na(age_tercile_2020)) %>%
  group_by(scenario, age_tercile_2020) %>%
  summarise(
    n                = n(),
    mean_DGHI        = mean(change_green, na.rm = TRUE),
    sd_DGHI          = sd(change_green, na.rm = TRUE),
    median_DGHI      = median(change_green, na.rm = TRUE),
    pct_brightened    = mean(change_green > 0, na.rm = TRUE) * 100,
    mean_mean_age     = mean(mean_age_2020, na.rm = TRUE),
    mean_NDVI_2010   = mean(NDVI_2010, na.rm = TRUE),
    mean_NDVI_2020   = mean(NDVI_2020, na.rm = TRUE),
    mean_delta_NDVI  = mean(delta_NDVI, na.rm = TRUE),
    .groups = "drop"
  )

cat("TABLE 8: ΔGHI by Age Composition Tercile (2020)\n")
print(as.data.frame(dghi_by_age), digits = 3, row.names = FALSE)
```

Same structure as Table 7, stratified by tract mean age.

**Read:** If the oldest tracts dim more, that indicates elderly populations are disproportionately exposed to greenness loss — relevant because older adults are more vulnerable to heat and benefit more from urban greenery.

```{r echo=FALSE}
# FIGURE 8: ΔGHI by age tercile (boxplot)
fig8 <- analysis_demo %>%
  filter(!is.na(age_tercile_2020)) %>%
  ggplot(aes(x = age_tercile_2020, y = change_green,
             fill = age_tercile_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = c("Youngest"     = "#deebf7",
                                "Middle-aged"  = "#9ecae1",
                                "Oldest"       = "#3182bd"),
                    guide = "none") +
  facet_wrap(~ scenario) +
  labs(title    = "Greenness Change by Tract Age Composition (2020)",
       subtitle = "Who is experiencing dimming vs. brightening?",
       x = "Tract age composition tercile (2020 mean age)",
       y = "ΔGHI (pseudo-residual change)") +
  theme_lambert

fig8
```

### Table 9 — Bright/Dim transitions by race tercile

For each race tercile and scenario, the four transition categories (Bright → Bright, Bright → Dim, Dim → Bright, Dim → Dim) are reported as counts and percentages.

```{r}
# --- TABLE 9: Bright/Dim transitions by race tercile ---
transitions_by_race <- analysis_demo %>%
  filter(!is.na(race_tercile_2020)) %>%
  group_by(scenario, race_tercile_2020, transition_green) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(scenario, race_tercile_2020) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

cat("TABLE 9: Bright/Dim Transitions by Race Tercile (2020)\n")
print(as.data.frame(transitions_by_race %>%
                      select(scenario, race_tercile_2020, transition_green, n, pct) %>%
                      arrange(scenario, race_tercile_2020, transition_green)),
      digits = 2, row.names = FALSE)
```

**Read:** This reveals whether the Bright → Dim asymmetry (dominant in the dormant season) is concentrated in particular racial composition groups. If high non-White tracts have a higher rate of Bright → Dim transitions, that is a stronger environmental justice signal than the mean ΔGHI alone.

### Table 10 — NDVI levels and housing growth by race tercile

```{r echo=FALSE}
# --- TABLE 10: Raw NDVI levels by race tercile (baseline exposure) ---
# This answers: do non-White tracts start with less greenness?
ndvi_levels_by_race <- analysis_demo %>%
  filter(!is.na(race_tercile_2020)) %>%
  group_by(scenario, race_tercile_2020) %>%
  summarise(
    n              = n(),
    mean_NDVI_2010 = mean(NDVI_2010, na.rm = TRUE),
    mean_NDVI_2020 = mean(NDVI_2020, na.rm = TRUE),
    mean_delta_NDVI = mean(delta_NDVI, na.rm = TRUE),
    mean_HU_dens_2010 = mean(HU_density_2010, na.rm = TRUE),
    mean_HU_dens_2020 = mean(HU_density_2020, na.rm = TRUE),
    mean_delta_HU  = mean(delta_HU, na.rm = TRUE),
    .groups = "drop"
  )

cat("TABLE 10: NDVI Levels & Housing by Race Tercile\n")
cat("  (Do non-White tracts start less green? Do they gain more housing?)\n")
print(as.data.frame(ndvi_levels_by_race), digits = 3, row.names = FALSE)
```


**Read:** This table connects the two guiding questions. It shows whether non-White tracts (a) start with less greenness, (b) lose more greenness, and (c) gain more housing. If high non-White tracts gain the most housing and lose the most greenness, that suggests housing growth in communities of color comes at a greenness cost.

### Table 11 — ΔGHI correlations with demographics and housing growth

Expanded correlation table, by scenario. Includes:
- Demographic *change* correlations (Δχ², Δ mean age, Δ % White, etc.)
- Demographic *level* correlations (2020 % non-White, 2020 mean age, 2020 % 65+)
- Housing growth correlations (ΔHU, ΔPOP)

```{r echo=FALSE}
# --- TABLE 11: ΔGHI correlations with demographic variables ---
# (Expanded version of the original Table 5, now including
# both levels and changes, and housing growth)
cor_by_scenario <- analysis_demo %>%
  group_by(scenario) %>%
  summarise(
    n = sum(!is.na(change_green)),
    # Demographic CHANGE correlations
    cor_dGHI_dchi_age       = cor(change_green, delta_chi_age,      use = "complete.obs"),
    cor_dGHI_dchi_race      = cor(change_green, delta_chi_race,     use = "complete.obs"),
    cor_dGHI_dmean_age      = cor(change_green, delta_mean_age,     use = "complete.obs"),
    cor_dGHI_dpct_under15   = cor(change_green, delta_pct_under15,  use = "complete.obs"),
    cor_dGHI_dpct_65p       = cor(change_green, delta_pct_65p,      use = "complete.obs"),
    cor_dGHI_dpct_white     = cor(change_green, delta_pct_white,    use = "complete.obs"),
    cor_dGHI_dpct_non_white = cor(change_green, delta_pct_non_white, use = "complete.obs"),
    # 2020 LEVEL correlations (who is exposed to the outcome)
    cor_dGHI_pctWhite2020   = cor(change_green, pct_white_2020,     use = "complete.obs"),
    cor_dGHI_pctNonWhite2020 = cor(change_green, pct_non_white_2020, use = "complete.obs"),
    cor_dGHI_meanAge2020    = cor(change_green, mean_age_2020,      use = "complete.obs"),
    cor_dGHI_pct65p2020     = cor(change_green, pct_65p_2020,       use = "complete.obs"),
    # Housing growth
    cor_dGHI_dHU            = cor(change_green, delta_HU,           use = "complete.obs"),
    cor_dGHI_dPOP           = cor(change_green, delta_POP,          use = "complete.obs"),
    .groups = "drop"
  )

# cat("TABLE 11: ΔGHI Correlations — Demographics & Housing Growth\n")
# print(as.data.frame(cor_by_scenario), digits = 3, row.names = FALSE)
```

| Variable | Dormant × 1000 m | Growing × 1000 m |
|---|---:|---:|
| **Demographic change** | | |
| Δχ² (Age) | -0.00547 | 0.11287 |
| Δχ² (Race) | 0.0319 | 0.1226 |
| Δ Mean age | 0.0296 | -0.1609 |
| Δ % Under 15 | -0.0952 | 0.0371 |
| Δ % Aged 65+ | -0.0869 | -0.1677 |
| Δ % White | -0.0247 | -0.0924 |
| Δ % Non-White | 0.0247 | 0.0924 |
| **2020 levels** | | |
| % White (2020) | -0.0258 | -0.0350 |
| % Non-White (2020) | 0.0258 | 0.0350 |
| Mean age (2020) | -0.0911 | -0.1579 |
| % Aged 65+ (2020) | -0.0686 | -0.1658 |
| **Housing growth** | | |
| ΔHU | 0.0443 | 0.2137 |
| ΔPOP | 0.0348 | 0.1557 |

**Read:** The level correlations are the key addition. The old analysis only included change correlations, which answer "do tracts that changed demographically also change in greenness?" The level correlations answer the more direct question: "are currently non-White tracts experiencing more dimming?" These are conceptually different — a tract can be stably non-White and still dim.

```{r echo=FALSE}
# FIGURE 12: Δ age variables vs Δχ²_age, faceted by scenario x y_var
age_long <- analysis_demo %>%
  select(tract, scenario, change_green,
         delta_chi_age,
         delta_mean_age, delta_pct_under15, delta_pct_65p) %>%
  pivot_longer(cols = c(delta_mean_age, delta_pct_under15, delta_pct_65p),
               names_to = "y_var", values_to = "y_value") %>%
  mutate(y_var = recode(y_var,
                        delta_mean_age    = "Δ Mean age (yrs)",
                        delta_pct_under15 = "Δ % Under 15",
                        delta_pct_65p     = "Δ % Aged 65+"))

fig12 <- ggplot(age_long,
               aes(x = delta_chi_age, y = y_value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = change_green), alpha = 0.6, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6) +
  scale_color_gradient2(low = "#878787", mid = "grey90", high = "#fdae61",
                        midpoint = 0, name = "ΔGHI") +
  facet_grid(y_var ~ scenario, scales = "free_y") +
  labs(title    = "Age Composition Change vs. Δχ² (Age)",
       subtitle = "x = chi-square distance between 2010 and 2020 age compositions",
       x = expression(Delta ~ chi[Age]^2),
       y = "Δ (age variable)") +
  theme_lambert

fig12

# FIGURE 13: Δ race variables vs Δχ²_race
race_long <- analysis_demo %>%
  select(tract, scenario, change_green,
         delta_chi_race,
         delta_pct_white, delta_pct_non_white) %>%
  pivot_longer(cols = c(delta_pct_white, delta_pct_non_white),
               names_to = "y_var", values_to = "y_value") %>%
  mutate(y_var = recode(y_var,
                        delta_pct_white     = "Δ % White",
                        delta_pct_non_white = "Δ % Non-White"))

fig13 <- ggplot(race_long,
               aes(x = delta_chi_race, y = y_value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = change_green), alpha = 0.6, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6) +
  scale_color_gradient2(low = "#878787", mid = "grey90", high = "#fdae61",
                        midpoint = 0, name = "ΔGHI") +
  facet_grid(y_var ~ scenario, scales = "free_y") +
  labs(title    = "Race Composition Change vs. Δχ² (Race)",
       subtitle = "x = chi-square distance between 2010 and 2020 race compositions",
       x = expression(Delta ~ chi[Race]^2),
       y = "Δ (race variable)") +
  theme_lambert

fig13

# FIGURE 14: ΔGHI vs Δχ² (greenness change vs demographic shift)
chi_long <- analysis_demo %>%
  select(tract, scenario, change_green, delta_chi_age, delta_chi_race) %>%
  pivot_longer(cols = c(delta_chi_age, delta_chi_race),
               names_to = "chi_var", values_to = "chi_value") %>%
  mutate(chi_var = recode(chi_var,
                          delta_chi_age  = "Δχ² (Age)",
                          delta_chi_race = "Δχ² (Race)"))

fig14 <- ggplot(chi_long,
               aes(x = chi_value, y = change_green)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.4, size = 0.7, color = "#2c7bb6") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6) +
  facet_grid(scenario ~ chi_var, scales = "free_x") +
  labs(title    = "ΔGHI vs. Demographic Δχ²",
       subtitle = "Does greenness change with demographic reshuffling?",
       x = "Δχ²", y = expression(Delta ~ "GHI")) +
  theme_lambert

fig14
```

### Table 12 — Demographic summary (all WA tracts)

```{r echo=FALSE}
# --- TABLE 12: Demographic summary (all tracts with demo data) ---
demo_summary <- demo %>%
  summarise(
    n_tracts               = n(),
    median_delta_chi_age   = median(delta_chi_age,   na.rm = TRUE),
    median_delta_chi_race  = median(delta_chi_race,  na.rm = TRUE),
    median_delta_mean_age  = median(delta_mean_age,  na.rm = TRUE),
    median_delta_pct_u15   = median(delta_pct_under15, na.rm = TRUE),
    median_delta_pct_65p   = median(delta_pct_65p,   na.rm = TRUE),
    median_delta_pct_white = median(delta_pct_white, na.rm = TRUE),
    mean_pct_white_2010    = mean(pct_white_2010, na.rm = TRUE),
    mean_pct_white_2020    = mean(pct_white_2020, na.rm = TRUE),
    mean_mean_age_2010     = mean(mean_age_2010, na.rm = TRUE),
    mean_mean_age_2020     = mean(mean_age_2020, na.rm = TRUE)
  )

# cat("TABLE 12: Demographic Summary (All WA Tracts)\n")
# print(as.data.frame(demo_summary), digits = 4, row.names = FALSE)
```

| Variable | Value |
|---|---:|
| n tracts | 1,784 |
| Median Δχ² (Age) | 0.057 |
| Median Δχ² (Race) | 0.072 |
| Median Δ mean age (years) | +1.79 |
| Median Δ % under 15 | −1.68 |
| Median Δ % aged 65+ | +3.90 |
| Median Δ % White | −7.73 |
| Mean % White (2010) | 82.63 |
| Mean % White (2020) | 73.87 |
| Mean age (2010) | 37.93 |
| Mean age (2020) | 39.63 |

**Read:** Washington tracts on average got older (mean age up ~1.8 years; share 65+ up ~3.9 points, share under 15 down ~1.7 points) and less White (−7.7 points).

---

## Interpretation

### What the previous version did well

The original tract analysis correctly established:
- The pseudo-residual framework works much better at the tract level (R² ~0.43–0.51 vs. ~0.01 at block level).
- Dormant season shows strong dimming; growing season shows mild brightening.
- The Bright → Dim asymmetry is sharper at the tract level than at the block level.
- Buffer is collapsible at the tract scale; season is not.

### What the previous version missed

The original analysis treated demographics as an afterthought — a correlation table (old Table 5) appended to the greenness analysis. That table showed weak correlations (|r| ≤ 0.17) between ΔGHI and demographic *change* variables, and concluded that "demographic change is weakly associated with greenness change." This framing had three problems:

1. **It answered the wrong question.** Correlating ΔGHI with *demographic change* asks "do tracts that changed demographically also change in greenness?" But the guiding question is "who is *experiencing* greenness change?" — which requires stratifying by demographic *levels*, not changes. A stably non-White tract that dims is an environmental justice concern regardless of whether its demographics shifted.

2. **It did not characterize housing growth.** The first guiding question asks "where is housing being built and who lives there?" The original analysis used housing density only as a regression predictor. It never summarized how much housing was added, where, or in what kinds of communities.

3. **It did not connect housing growth to demographics to greenness.** The three-way relationship — housing is built in tract X, which has demographic profile Y, and experiences greenness change Z — was never examined.

### What the updated analysis adds

1. **Housing growth characterization (Tables 4–6, Figure 11).** We now know how many housing units were added statewide, which tracts gained the most, and the demographic profile of high-growth vs. low-growth tracts. This directly answers "where is housing being built and who lives there?"

2. **Greenness exposure by demographic group (Tables 7–10, Figures 7–10).** By stratifying ΔGHI by race and age terciles, we directly answer "who is experiencing changes in green exposure?" If high non-White tracts dim more in the dormant season, that is a finding about environmental justice — not just a correlation.

3. **Baseline exposure disparities (Table 10, Figure 10).** We now report whether non-White tracts start with less greenness in 2010. If they do, and they also dim more, that is a compounding disadvantage.

4. **Transition asymmetries by race (Table 9).** The Bright → Dim asymmetry is the most striking finding of the dormant-season analysis. Knowing whether it falls disproportionately on non-White tracts is essential for the environmental justice framing.

5. **Level vs. change correlations (Table 11).** The expanded correlation table separates demographic *levels* (who lives there in 2020) from demographic *changes* (how the tract shifted). This distinction matters because the policy-relevant question is about current exposure, not about whether demographic change tracks greenness change.

### Ecological interpretation (unchanged from previous version)

- Dormant-season dimming suggests loss of evergreen / structural cover — the vegetation visible during leaf-off conditions.
- Growing-season brightening suggests active-season canopy held or improved.
- A tract can simultaneously lose winter greenness and gain summer greenness. This is a feature of the data, not a contradiction.

---

## Figures generated (`output_updated/`)

### Greenness analysis (same as before)
1. `tract_fig1_baseline_green_regression.png` — 2010 NDVI vs. log(HU density), 2 scenarios.
2. `tract_fig2_change_green_vs_housing.png` — ΔGHI vs. Δlog(HU density).
3. `tract_fig3_change_green_vs_ndvi.png` — ΔGHI vs. raw ΔNDVI.
4. `tract_fig4_arrow_green_trajectories.png` — Arrow plot of 2010 → 2020 trajectories.
5. `tract_fig5_boxplot_change.png` — ΔGHI distribution by scenario.
6. `tract_fig6_mean_change_ci.png` — Mean ΔGHI with 95% CI per scenario.

### Demographic stratification (new)
7. `tract_fig7_dghi_by_race.png` — **ΔGHI boxplots by race composition tercile.** Answers: do non-White tracts dim more?
8. `tract_fig8_dghi_by_age.png` — **ΔGHI boxplots by age composition tercile.** Answers: do older tracts dim more?
9. `tract_fig9_hu_growth_dghi_by_race.png` — **Housing growth vs. ΔGHI, colored by race tercile.** Connects housing growth, greenness change, and race.
10. `tract_fig10_baseline_ndvi_by_race.png` — **Baseline NDVI (2010) by race tercile.** Answers: do non-White tracts start less green?
11. `tract_fig11_demo_by_hu_growth.png` — **Demographics of housing growth quartiles.** Answers: who lives where housing is being built?

### Demographic change (carried forward, renumbered)
12. `tract_fig12_age_vs_delta_chi.png` — Δ(age variable) vs. Δχ²(Age); facets: scenario × y variable.
13. `tract_fig13_race_vs_delta_chi.png` — Δ(race variable) vs. Δχ²(Race); facets: scenario × y variable.
14. `tract_fig14_dghi_vs_delta_chi.png` — ΔGHI vs. Δχ² (age and race) by scenario.

## Tables generated (`output_updated/`)

### Greenness analysis
- `tract_table1_regression_coefficients.csv`
- `tract_table2_change_summary.csv`
- `tract_table3_transitions.csv`

### Housing growth (new)
- `tract_table4_housing_summary.csv` — Statewide housing growth summary
- `tract_table5_demo_by_hu_growth.csv` — Demographics by housing growth quartile
- `tract_table6_hu_demo_correlations.csv` — Housing growth × demographic correlations

### Greenness × demographics (new)
- `tract_table7_dghi_by_race.csv` — ΔGHI by race composition tercile
- `tract_table8_dghi_by_age.csv` — ΔGHI by age composition tercile
- `tract_table9_transitions_by_race.csv` — Bright/Dim transitions by race tercile
- `tract_table10_ndvi_levels_by_race.csv` — NDVI levels and housing by race tercile
- `tract_table11_dghi_correlations.csv` — Expanded ΔGHI correlation table
- `tract_table12_demographic_summary.csv` — Demographic summary (all WA tracts)

### Data files
- `tract_demographics_with_delta_chi.csv` — Per-tract demographic deltas including Δχ²
- `tract_pseudo_residual_analysis.csv` — Full analytical dataset (greenness + demographics)

---

## Reference

Lambert, M. R., Des Roches, S., Auerbach, D. A., Van Deynze, B., Behrens, S., Hale, R., & Pierce, K. B. (2025). Building the neighborhood for the trees: Illuminating win–wins for housing densification and nature. *Conservation Science and Practice*, 7(7), e70085. https://doi.org/10.1111/csp2.70085
