---
title: "Tract-Level Analysis: Housing Growth, Demographic Change, and Greenness Exposure"
author: "Lilac Hong"
date: "2026-04-30"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: true
---

```{r include = FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)

here::i_am("code/2_2_tract_ndvi.Rmd")
```

## Guiding questions

1. **Where is housing being built and who lives there** (and how have demographics changed over time)?
2. **Who is experiencing changes in exposure to green environments?**

## Demographic categories

- **Race:** AIAN, Asian, Black, NHOPI, Two or More, White (individual categories, not collapsed)
- **Age:** 0–19, 20–29, 30–59, 60+

---

## Why only 1000 m at the tract level?

Buffer main effect is not statistically significant (p = 0.10 dormant, p = 0.39 growing). Cross-buffer ΔGHI correlations are 0.75–0.99. Cohen's d for 500 vs. 2000 m is negligible (|d| ≤ 0.07). AIC/BIC prefer the season-only model. The choice of 1000 m is arbitrary among the three — any buffer gives the same answer.

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# ============================================================
# DATA PREPARATION (sections 1-4 from 2_2_tract_ndvi.R)
# ============================================================
ndvi_tract <- read.csv(here::here("data", "tract_final.csv"))

df <- ndvi_tract %>%
  mutate(
    across(all_of(c("NDVI", "HU_density")), as.numeric),
    year   = as.integer(year),
    buffer = as.integer(buffer)
  ) %>%
  filter(year %in% c(2010, 2020)) %>%
  filter(!is.na(NDVI), !is.na(HU_density), HU_density > 0) %>%
  mutate(log_HU_density = log(HU_density))

wide <- df %>%
  select(tract, season, buffer, year, log_HU_density, NDVI, HU, POP, area_ha) %>%
  pivot_wider(names_from  = year,
              values_from = c(log_HU_density, NDVI, HU, POP),
              names_sep   = "_") %>%
  drop_na() %>%
  mutate(delta_log_HU = log_HU_density_2020 - log_HU_density_2010,
         delta_NDVI   = NDVI_2020 - NDVI_2010,
         delta_HU     = HU_2020 - HU_2010,
         delta_POP    = POP_2020 - POP_2010,
         HU_density_2010 = HU_2010 / area_ha,
         HU_density_2020 = HU_2020 / area_ha,
         delta_HU_density = HU_density_2020 - HU_density_2010)

scenarios <- tibble::tribble(
  ~season,           ~buffer,
  "Dormant_Season",     1000,
  "Growing_Season",     1000
)

wide <- wide %>% semi_join(scenarios, by = c("season", "buffer"))

theme_lambert <- theme_minimal(base_size = 11) +
  theme(strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        legend.position  = "bottom")
```

## Part I: Pseudo-Residual Greenness Analysis

### Methods

$\text{NDVI}_{2010} = \alpha + \beta \cdot \log(\text{HU_density}_{2010})$

- GHI~2010~ = NDVI~2010~ − ($\hat{\alpha}$ + $\hat{\beta}$ · log(HU_density~2010~))
- GHI~2020~ = NDVI~2020~ − ($\hat{\alpha}$ + $\hat{\beta}$ · log(HU_density~2020~))
- **ΔGHI = GHI~2020~ − GHI~2010~**

Sample: **`r nrow(wide)`** tract × scenario rows across **`r n_distinct(wide$tract)`** unique tracts.

### Table 1 — 2010 baseline regressions

```{r echo=FALSE, message=FALSE}
results <- scenarios %>%
  mutate(data = map2(season, buffer, ~ wide %>%
                       filter(season == .x, buffer == .y))) %>%
  mutate(mod_green = map(data, ~ lm(NDVI_2010 ~ log_HU_density_2010, data = .x)))

reg_table <- results %>%
  mutate(
    tidy_green = map(mod_green, ~ {
      s <- summary(.x)
      tibble(intercept = coef(.x)[1], slope = coef(.x)[2],
             slope_se = coef(s)[2, 2], slope_p = coef(s)[2, 4],
             R2 = s$r.squared, n = nobs(.x))
    })
  ) %>%
  select(season, buffer, tidy_green) %>%
  unnest(tidy_green)

knitr::kable(reg_table, digits = 4)
```

### Table 2 — ΔGHI summary

```{r echo=FALSE, message=FALSE}
results <- results %>%
  mutate(
    data = map2(data, mod_green, function(d, mg) {
      d %>% mutate(
        resid_green_2010 = NDVI_2010 - predict(mg, newdata = d),
        pred_green_2020  = predict(mg, newdata = data.frame(
          log_HU_density_2010 = d$log_HU_density_2020)),
        resid_green_2020 = NDVI_2020 - pred_green_2020,
        change_green     = resid_green_2020 - resid_green_2010)
    })
  )

all_data <- results %>%
  select(data) %>% unnest(data) %>%
  mutate(
    state_green_2010 = ifelse(resid_green_2010 >= 0, "Bright", "Dim"),
    state_green_2020 = ifelse(resid_green_2020 >= 0, "Bright", "Dim"),
    transition_green = paste0(state_green_2010, " -> ", state_green_2020),
    direction_green  = case_when(
      change_green > 0 ~ "Brightened",
      change_green < 0 ~ "Dimmed",
      TRUE ~ "No change"),
    scenario = factor(paste(ifelse(season == "Dormant_Season", "Dormant", "Growing"), "1000 m"),
                      levels = c("Dormant 1000 m", "Growing 1000 m"))
  )

change_summary <- all_data %>%
  group_by(season, buffer) %>%
  summarise(n = n(), mean = mean(change_green), sd = sd(change_green),
            median = median(change_green),
            pct_bright = mean(change_green > 0) * 100,
            pct_dim = mean(change_green < 0) * 100, .groups = "drop")

knitr::kable(change_summary, digits = 4)
```

### Table 3 — Bright/Dim transitions

```{r echo=FALSE}
trans_tbl <- all_data %>%
  group_by(season, buffer, transition_green) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = transition_green, values_from = n, values_fill = 0)

knitr::kable(trans_tbl)
```

### Paired t-test — Dormant vs. Growing

```{r echo=FALSE}
paired <- all_data %>%
  select(tract, season, change_green) %>%
  pivot_wider(names_from = season, values_from = change_green) %>%
  drop_na()
t.test(paired$Dormant_Season, paired$Growing_Season, paired = TRUE)
```


### Greenness figures

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Fig 1: 2010 baseline
ggplot(all_data, aes(x = log_HU_density_2010, y = NDVI_2010)) +
  geom_point(alpha = 0.2, size = 0.5, color = "grey40") +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6", linewidth = 0.8) +
  facet_wrap(~ scenario) +
  labs(title = "2010 Baseline: Greenness vs. Housing Density",
       x = "log(HU density/ha), 2010", y = "NDVI, 2010") +
  theme_lambert
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Fig 2: ΔGHI vs housing density change
ggplot(all_data, aes(x = delta_log_HU, y = change_green)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_point(aes(color = direction_green), alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.7) +
  scale_color_manual(values = c("Brightened" = "#fdae61", "Dimmed" = "#878787",
                                "No change" = "grey50"), name = "Direction") +
  facet_wrap(~ scenario) +
  labs(title = "ΔGHI vs. Housing Density Change",
       x = expression(Delta ~ "log(HU density/ha)"), y = expression(Delta ~ "GHI")) +
  theme_lambert
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Fig 3: ΔGHI vs raw NDVI change
ggplot(all_data, aes(x = delta_NDVI, y = change_green)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_point(aes(color = direction_green), alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.7) +
  scale_color_manual(values = c("Brightened" = "#fdae61", "Dimmed" = "#878787",
                                "No change" = "grey50"), name = "Direction") +
  facet_wrap(~ scenario) +
  labs(title = "ΔGHI vs. Raw NDVI Change",
       x = expression(Delta ~ "NDVI"), y = expression(Delta ~ "GHI")) +
  theme_lambert
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
# Fig 5: Boxplot
ggplot(all_data, aes(x = scenario, y = change_green, fill = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.5, width = 0.5) +
  scale_fill_manual(values = c("Dormant 1000 m" = "#abd9e9",
                                "Growing 1000 m" = "#66c164"), guide = "none") +
  labs(title = "Distribution of ΔGHI by Scenario", x = NULL, y = "ΔGHI") +
  theme_lambert
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
# Fig 6: Mean ΔGHI with 95% CI
ci_data <- all_data %>%
  group_by(scenario) %>%
  summarise(mean = mean(change_green), se = sd(change_green) / sqrt(n()), .groups = "drop") %>%
  mutate(lo = mean - 1.96 * se, hi = mean + 1.96 * se)

ggplot(ci_data, aes(x = scenario, y = mean, color = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.8) +
  scale_color_manual(values = c("Dormant 1000 m" = "#2c7bb6",
                                 "Growing 1000 m" = "#1a9641"), guide = "none") +
  labs(title = "Mean ΔGHI with 95% CI", x = NULL, y = "Mean ΔGHI") +
  theme_lambert
```

---


## Part II: Where Is Housing Being Built and Who Lives There?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# ============================================================
# DEMOGRAPHICS: 6 race categories, 4 age groups
# ============================================================
age  <- read.csv(here::here("data", "age", "age_10_20.csv"))
race <- read.csv(here::here("data", "race","race_10_20.csv"))

age$tract  <- as.numeric(gsub("[^0-9]", "", as.character(age$tract20l)))
race$tract <- as.numeric(gsub("[^0-9]", "", as.character(race$tract20l)))

age_bands <- c("age0_4","age5_9","age10_14","age15_19","age20_24","age25_29",
               "age30_34","age35_39","age40_44","age45_49","age50_54","age55_59",
               "age60_64","age65_69","age70_74","age75_79","age80_84","age85_up")

age_group_pct <- function(df, yr) {
  cols <- paste0(age_bands, "_", yr)
  A <- as.matrix(df[, cols])
  total <- rowSums(A)
  total_safe <- ifelse(total == 0, NA_real_, total)
  data.frame(total = total,
             pct_0_19  = rowSums(A[, 1:4])   / total_safe * 100,
             pct_20_29 = rowSums(A[, 5:6])   / total_safe * 100,
             pct_30_59 = rowSums(A[, 7:12])  / total_safe * 100,
             pct_60p   = rowSums(A[, 13:18]) / total_safe * 100)
}
a10 <- age_group_pct(age, 2010)
a20 <- age_group_pct(age, 2020)

age_delta_chi <- function(df) {
  A10 <- as.matrix(df[, paste0(age_bands, "_2010")])
  A20 <- as.matrix(df[, paste0(age_bands, "_2020")])
  G10 <- cbind(rowSums(A10[,1:4]), rowSums(A10[,5:6]),
               rowSums(A10[,7:12]), rowSums(A10[,13:18]))
  G20 <- cbind(rowSums(A20[,1:4]), rowSums(A20[,5:6]),
               rowSums(A20[,7:12]), rowSums(A20[,13:18]))
  T10 <- rowSums(G10); T20 <- rowSums(G20)
  P10 <- G10 / ifelse(T10 == 0, NA_real_, T10)
  P20 <- G20 / ifelse(T20 == 0, NA_real_, T20)
  Pbar <- (P10 + P20) / 2
  num  <- (P20 - P10)^2
  ratio <- ifelse(Pbar > 0, num / Pbar, 0)
  rowSums(ratio, na.rm = TRUE)
}

age_out <- data.frame(
  tract = age$tract, total_pop_2010 = a10$total, total_pop_2020 = a20$total,
  pct_0_19_2010 = a10$pct_0_19, pct_0_19_2020 = a20$pct_0_19,
  pct_20_29_2010 = a10$pct_20_29, pct_20_29_2020 = a20$pct_20_29,
  pct_30_59_2010 = a10$pct_30_59, pct_30_59_2020 = a20$pct_30_59,
  pct_60p_2010 = a10$pct_60p, pct_60p_2020 = a20$pct_60p,
  delta_chi_age = age_delta_chi(age)
)
age_out$delta_pct_0_19  <- age_out$pct_0_19_2020  - age_out$pct_0_19_2010
age_out$delta_pct_20_29 <- age_out$pct_20_29_2020 - age_out$pct_20_29_2010
age_out$delta_pct_30_59 <- age_out$pct_30_59_2020 - age_out$pct_30_59_2010
age_out$delta_pct_60p   <- age_out$pct_60p_2020   - age_out$pct_60p_2010
age_out$delta_pop       <- age_out$total_pop_2020  - age_out$total_pop_2010

# Race
race_cats   <- c("AIAN", "ASIAN", "Black", "NHOPI", "Two", "White")
race_labels <- c("AIAN", "Asian", "Black", "NHOPI", "Two_or_More", "White")

R10 <- as.matrix(race[, paste0(race_cats, "2010")])
R20 <- as.matrix(race[, paste0(race_cats, "2020")])
T10 <- rowSums(R10); T20 <- rowSums(R20)
P10 <- R10 / ifelse(T10 == 0, NA_real_, T10)
P20 <- R20 / ifelse(T20 == 0, NA_real_, T20)
Pbar_race <- (P10 + P20) / 2
num_race  <- (P20 - P10)^2
ratio_race <- ifelse(Pbar_race > 0, num_race / Pbar_race, 0)

race_out <- data.frame(
  tract = race$tract, total_race_2010 = T10, total_race_2020 = T20,
  delta_chi_race = rowSums(ratio_race, na.rm = TRUE)
)
for (i in seq_along(race_cats)) {
  race_out[[paste0("pct_", race_labels[i], "_2010")]] <- P10[, paste0(race_cats[i], "2010")] * 100
  race_out[[paste0("pct_", race_labels[i], "_2020")]] <- P20[, paste0(race_cats[i], "2020")] * 100
}
for (rl in race_labels) {
  race_out[[paste0("delta_pct_", rl)]] <-
    race_out[[paste0("pct_", rl, "_2020")]] - race_out[[paste0("pct_", rl, "_2010")]]
}

demo <- age_out %>% full_join(race_out, by = "tract")
analysis_demo <- all_data %>% left_join(demo, by = "tract")
```

### Table 4 — Housing growth summary

```{r echo=FALSE}
tract_housing <- analysis_demo %>%
  filter(season == "Dormant_Season") %>%
  select(tract, area_ha, HU_2010, HU_2020, POP_2010, POP_2020,
         delta_HU, delta_POP, delta_HU_density,
         HU_density_2010, HU_density_2020,
         starts_with("pct_"), starts_with("delta_"),
         starts_with("total_pop"), starts_with("total_race"))

housing_summary <- tract_housing %>%
  summarise(
    n_tracts = n(),
    total_HU_2010 = sum(HU_2010, na.rm = TRUE),
    total_HU_2020 = sum(HU_2020, na.rm = TRUE),
    total_delta_HU = sum(delta_HU, na.rm = TRUE),
    pct_gained = mean(delta_HU > 0, na.rm = TRUE) * 100,
    pct_lost = mean(delta_HU < 0, na.rm = TRUE) * 100,
    median_delta_HU = median(delta_HU, na.rm = TRUE),
    mean_delta_HU = mean(delta_HU, na.rm = TRUE),
    total_POP_2010 = sum(POP_2010, na.rm = TRUE),
    total_POP_2020 = sum(POP_2020, na.rm = TRUE),
    total_delta_POP = sum(delta_POP, na.rm = TRUE)
  )

knitr::kable(t(housing_summary), col.names = "Value", digits = 1)
```

### Table 5 — Demographics by housing growth quartile

```{r echo=FALSE}
tract_housing <- tract_housing %>%
  mutate(hu_growth_q = cut(delta_HU,
    breaks = quantile(delta_HU, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE),
    labels = c("Q1 (least)", "Q2", "Q3", "Q4 (most)"),
    include.lowest = TRUE))

demo_by_hu <- tract_housing %>%
  group_by(hu_growth_q) %>%
  summarise(
    n = n(), mean_dHU = mean(delta_HU, na.rm = TRUE),
    # Race 2020
    across(all_of(paste0("pct_", race_labels, "_2020")),
           ~ mean(.x, na.rm = TRUE), .names = "mean_{.col}"),
    # Age 2020
    mean_pct_0_19_2020 = mean(pct_0_19_2020, na.rm = TRUE),
    mean_pct_20_29_2020 = mean(pct_20_29_2020, na.rm = TRUE),
    mean_pct_30_59_2020 = mean(pct_30_59_2020, na.rm = TRUE),
    mean_pct_60p_2020 = mean(pct_60p_2020, na.rm = TRUE),
    # Race 2010
    across(all_of(paste0("pct_", race_labels, "_2010")),
           ~ mean(.x, na.rm = TRUE), .names = "mean_{.col}"),
    # Age 2010
    mean_pct_0_19_2010 = mean(pct_0_19_2010, na.rm = TRUE),
    mean_pct_20_29_2010 = mean(pct_20_29_2010, na.rm = TRUE),
    mean_pct_30_59_2010 = mean(pct_30_59_2010, na.rm = TRUE),
    mean_pct_60p_2010 = mean(pct_60p_2010, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(demo_by_hu, digits = 1)
```

### Figure — Race composition by housing growth quartile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
race_by_hu_long <- demo_by_hu %>%
  select(hu_growth_q,
         starts_with("mean_pct_") & matches("(AIAN|Asian|Black|NHOPI|Two_or_More|White)")) %>%
  pivot_longer(-hu_growth_q, names_to = "variable", values_to = "value") %>%
  mutate(
    year = ifelse(grepl("2010", variable), "2010", "2020"),
    race = gsub("mean_pct_(.+)_(2010|2020)", "\\1", variable),
    race = gsub("_", " ", race))

ggplot(race_by_hu_long, aes(x = hu_growth_q, y = value, fill = race)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_brewer(palette = "Set2", name = "Race") +
  facet_wrap(~ year) +
  labs(title = "Race Composition of Housing Growth Quartiles",
       x = "Housing growth quartile (ΔHU)", y = "Mean % of tract population") +
  theme_lambert + theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

### Figure — Age composition by housing growth quartile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
age_by_hu_long <- demo_by_hu %>%
  select(hu_growth_q, matches("mean_pct_(0_19|20_29|30_59|60p)_(2010|2020)")) %>%
  pivot_longer(-hu_growth_q, names_to = "variable", values_to = "value") %>%
  mutate(
    year = ifelse(grepl("2010", variable), "2010", "2020"),
    age_group = case_when(
      grepl("0_19", variable)  ~ "0-19",
      grepl("20_29", variable) ~ "20-29",
      grepl("30_59", variable) ~ "30-59",
      grepl("60p", variable)   ~ "60+"),
    age_group = factor(age_group, levels = c("0-19", "20-29", "30-59", "60+")))

ggplot(age_by_hu_long, aes(x = hu_growth_q, y = value, fill = age_group)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_brewer(palette = "Blues", name = "Age group") +
  facet_wrap(~ year) +
  labs(title = "Age Composition of Housing Growth Quartiles",
       x = "Housing growth quartile (ΔHU)", y = "Mean % of tract population") +
  theme_lambert + theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

### Table 6 — Correlations: housing growth × demographics

```{r echo=FALSE}
race_pct_2010_cols <- paste0("pct_", race_labels, "_2010")
race_pct_2020_cols <- paste0("pct_", race_labels, "_2020")
race_delta_cols    <- paste0("delta_pct_", race_labels)

cor_vars <- c(race_pct_2010_cols, race_pct_2020_cols, race_delta_cols,
              "pct_0_19_2010","pct_20_29_2010","pct_30_59_2010","pct_60p_2010",
              "pct_0_19_2020","pct_20_29_2020","pct_30_59_2020","pct_60p_2020",
              "delta_pct_0_19","delta_pct_20_29","delta_pct_30_59","delta_pct_60p",
              "delta_chi_age","delta_chi_race")

hu_cors <- sapply(cor_vars, function(v) {
  cor(tract_housing$delta_HU, tract_housing[[v]], use = "complete.obs")
})

knitr::kable(data.frame(variable = names(hu_cors), r_dHU = round(hu_cors, 3)),
             row.names = FALSE)
```

---


## Part III: Who Is Experiencing Changes in Exposure to Green Environments?

```{r echo=FALSE}
analysis_demo <- analysis_demo %>%
  mutate(
    white_tercile_2020 = cut(pct_White_2020,
      breaks = quantile(pct_White_2020, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      labels = c("Low White", "Mid White", "High White"),
      include.lowest = TRUE),
    pct_60p_tercile_2020 = cut(pct_60p_2020,
      breaks = quantile(pct_60p_2020, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      labels = c("Low 60+", "Mid 60+", "High 60+"),
      include.lowest = TRUE)
  )
```

### Table 7 — ΔGHI by % White tercile (2020)

```{r echo=FALSE}
dghi_by_white <- analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  group_by(scenario, white_tercile_2020) %>%
  summarise(n = n(), mean_DGHI = mean(change_green, na.rm = TRUE),
            sd = sd(change_green, na.rm = TRUE),
            median = median(change_green, na.rm = TRUE),
            pct_bright = mean(change_green > 0, na.rm = TRUE) * 100,
            mean_pct_White = mean(pct_White_2020, na.rm = TRUE),
            mean_NDVI_2010 = mean(NDVI_2010, na.rm = TRUE),
            mean_NDVI_2020 = mean(NDVI_2020, na.rm = TRUE),
            mean_dNDVI = mean(delta_NDVI, na.rm = TRUE),
            .groups = "drop")

knitr::kable(dghi_by_white, digits = 3)
```

### Figure — ΔGHI by % White tercile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  ggplot(aes(x = white_tercile_2020, y = change_green, fill = white_tercile_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = c("Low White" = "#e34a33", "Mid White" = "#fdbb84",
                                "High White" = "#fee8c8"), guide = "none") +
  facet_wrap(~ scenario) +
  labs(title = "Greenness Change by Tract % White (2020)",
       x = "% White tercile (2020)", y = "ΔGHI") +
  theme_lambert
```

### Table 8 — ΔGHI by % 60+ tercile (2020)

```{r echo=FALSE}
dghi_by_60p <- analysis_demo %>%
  filter(!is.na(pct_60p_tercile_2020)) %>%
  group_by(scenario, pct_60p_tercile_2020) %>%
  summarise(n = n(), mean_DGHI = mean(change_green, na.rm = TRUE),
            sd = sd(change_green, na.rm = TRUE),
            median = median(change_green, na.rm = TRUE),
            pct_bright = mean(change_green > 0, na.rm = TRUE) * 100,
            mean_pct_60p = mean(pct_60p_2020, na.rm = TRUE),
            mean_NDVI_2010 = mean(NDVI_2010, na.rm = TRUE),
            mean_NDVI_2020 = mean(NDVI_2020, na.rm = TRUE),
            mean_dNDVI = mean(delta_NDVI, na.rm = TRUE),
            .groups = "drop")

knitr::kable(dghi_by_60p, digits = 3)
```

### Figure — ΔGHI by % 60+ tercile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
analysis_demo %>%
  filter(!is.na(pct_60p_tercile_2020)) %>%
  ggplot(aes(x = pct_60p_tercile_2020, y = change_green, fill = pct_60p_tercile_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = c("Low 60+" = "#deebf7", "Mid 60+" = "#9ecae1",
                                "High 60+" = "#3182bd"), guide = "none") +
  facet_wrap(~ scenario) +
  labs(title = "Greenness Change by Tract % Aged 60+ (2020)",
       x = "% 60+ tercile (2020)", y = "ΔGHI") +
  theme_lambert
```

### Table 9 — Bright/Dim transitions by % White tercile

```{r echo=FALSE}
trans_by_white <- analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  group_by(scenario, white_tercile_2020, transition_green) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(scenario, white_tercile_2020) %>%
  mutate(pct = n / sum(n) * 100) %>% ungroup()

knitr::kable(trans_by_white %>%
  select(scenario, white_tercile_2020, transition_green, n, pct) %>%
  arrange(scenario, white_tercile_2020, transition_green), digits = 1)
```

### Table 10 — NDVI levels & housing by % White tercile

```{r echo=FALSE}
ndvi_by_white <- analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  group_by(scenario, white_tercile_2020) %>%
  summarise(n = n(),
            mean_NDVI_2010 = mean(NDVI_2010, na.rm = TRUE),
            mean_NDVI_2020 = mean(NDVI_2020, na.rm = TRUE),
            mean_dNDVI = mean(delta_NDVI, na.rm = TRUE),
            mean_HU_dens_2010 = mean(HU_density_2010, na.rm = TRUE),
            mean_HU_dens_2020 = mean(HU_density_2020, na.rm = TRUE),
            mean_dHU = mean(delta_HU, na.rm = TRUE),
            .groups = "drop")

knitr::kable(ndvi_by_white, digits = 3)
```

### Figure — Baseline NDVI (2010) by % White tercile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  ggplot(aes(x = white_tercile_2020, y = NDVI_2010, fill = white_tercile_2020)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, width = 0.6) +
  scale_fill_manual(values = c("Low White" = "#e34a33", "Mid White" = "#fdbb84",
                                "High White" = "#fee8c8"), guide = "none") +
  facet_wrap(~ scenario) +
  labs(title = "Baseline Greenness (2010) by Tract % White",
       x = "% White tercile (2020)", y = "NDVI, 2010") +
  theme_lambert
```

### Figure — Housing growth vs. ΔGHI by % White tercile

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
analysis_demo %>%
  filter(!is.na(white_tercile_2020)) %>%
  ggplot(aes(x = delta_HU, y = change_green, color = white_tercile_2020)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_color_manual(values = c("Low White" = "#e34a33", "Mid White" = "#fdbb84",
                                 "High White" = "#fee8c8"),
                     name = "% White tercile") +
  facet_wrap(~ scenario) +
  labs(title = "Housing Growth vs. Greenness Change, by % White",
       x = "ΔHU (2010-2020)", y = "ΔGHI") +
  theme_lambert
```


### Table 11 — ΔGHI correlations with all demographic variables

```{r echo=FALSE}
cor_by_scenario <- analysis_demo %>%
  group_by(scenario) %>%
  summarise(
    n = sum(!is.na(change_green)),
    across(all_of(paste0("pct_", race_labels, "_2020")),
           ~ cor(change_green, .x, use = "complete.obs"), .names = "cor_dGHI_{.col}"),
    across(all_of(paste0("delta_pct_", race_labels)),
           ~ cor(change_green, .x, use = "complete.obs"), .names = "cor_dGHI_{.col}"),
    cor_dGHI_pct_0_19_2020  = cor(change_green, pct_0_19_2020, use = "complete.obs"),
    cor_dGHI_pct_20_29_2020 = cor(change_green, pct_20_29_2020, use = "complete.obs"),
    cor_dGHI_pct_30_59_2020 = cor(change_green, pct_30_59_2020, use = "complete.obs"),
    cor_dGHI_pct_60p_2020   = cor(change_green, pct_60p_2020, use = "complete.obs"),
    cor_dGHI_delta_pct_0_19  = cor(change_green, delta_pct_0_19, use = "complete.obs"),
    cor_dGHI_delta_pct_20_29 = cor(change_green, delta_pct_20_29, use = "complete.obs"),
    cor_dGHI_delta_pct_30_59 = cor(change_green, delta_pct_30_59, use = "complete.obs"),
    cor_dGHI_delta_pct_60p   = cor(change_green, delta_pct_60p, use = "complete.obs"),
    cor_dGHI_dchi_age  = cor(change_green, delta_chi_age, use = "complete.obs"),
    cor_dGHI_dchi_race = cor(change_green, delta_chi_race, use = "complete.obs"),
    cor_dGHI_dHU  = cor(change_green, delta_HU, use = "complete.obs"),
    cor_dGHI_dPOP = cor(change_green, delta_POP, use = "complete.obs"),
    .groups = "drop")

# Transpose for readability
cor_long <- cor_by_scenario %>%
  select(-n) %>%
  pivot_longer(-scenario, names_to = "variable", values_to = "r") %>%
  pivot_wider(names_from = scenario, values_from = r)

knitr::kable(cor_long, digits = 3)
```

### Table 12 — Demographic summary (all WA tracts)

```{r echo=FALSE}
demo_summary <- demo %>%
  summarise(
    n_tracts = n(),
    across(all_of(c(paste0("pct_", race_labels, "_2010"),
                    paste0("pct_", race_labels, "_2020"))),
           ~ mean(.x, na.rm = TRUE), .names = "mean_{.col}"),
    mean_pct_0_19_2010 = mean(pct_0_19_2010, na.rm = TRUE),
    mean_pct_0_19_2020 = mean(pct_0_19_2020, na.rm = TRUE),
    mean_pct_20_29_2010 = mean(pct_20_29_2010, na.rm = TRUE),
    mean_pct_20_29_2020 = mean(pct_20_29_2020, na.rm = TRUE),
    mean_pct_30_59_2010 = mean(pct_30_59_2010, na.rm = TRUE),
    mean_pct_30_59_2020 = mean(pct_30_59_2020, na.rm = TRUE),
    mean_pct_60p_2010 = mean(pct_60p_2010, na.rm = TRUE),
    mean_pct_60p_2020 = mean(pct_60p_2020, na.rm = TRUE),
    median_delta_chi_age = median(delta_chi_age, na.rm = TRUE),
    median_delta_chi_race = median(delta_chi_race, na.rm = TRUE)
  )

knitr::kable(t(demo_summary), col.names = "Value", digits = 2)
```

### Figure — Race composition change vs. Δχ² (Race)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=12}
race_delta_long <- analysis_demo %>%
  select(tract, scenario, change_green, delta_chi_race,
         all_of(paste0("delta_pct_", race_labels))) %>%
  pivot_longer(cols = starts_with("delta_pct_"),
               names_to = "y_var", values_to = "y_value") %>%
  mutate(y_var = gsub("delta_pct_", "Δ % ", y_var),
         y_var = gsub("_", " ", y_var))

ggplot(race_delta_long, aes(x = delta_chi_race, y = y_value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = change_green), alpha = 0.5, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
  scale_color_gradient2(low = "#878787", mid = "grey90", high = "#fdae61",
                        midpoint = 0, name = "ΔGHI") +
  facet_grid(y_var ~ scenario, scales = "free_y") +
  labs(title = "Race Composition Change vs. Δχ² (Race)",
       x = expression(Delta ~ chi[Race]^2), y = "Δ (race %)") +
  theme_lambert + theme(strip.text.y = element_text(size = 8))
```

### Figure — Age composition change vs. Δχ² (Age)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
age_delta_long <- analysis_demo %>%
  select(tract, scenario, change_green, delta_chi_age,
         delta_pct_0_19, delta_pct_20_29, delta_pct_30_59, delta_pct_60p) %>%
  pivot_longer(cols = starts_with("delta_pct_"),
               names_to = "y_var", values_to = "y_value") %>%
  mutate(y_var = recode(y_var,
    delta_pct_0_19 = "Δ % 0-19", delta_pct_20_29 = "Δ % 20-29",
    delta_pct_30_59 = "Δ % 30-59", delta_pct_60p = "Δ % 60+"))

ggplot(age_delta_long, aes(x = delta_chi_age, y = y_value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = change_green), alpha = 0.5, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
  scale_color_gradient2(low = "#878787", mid = "grey90", high = "#fdae61",
                        midpoint = 0, name = "ΔGHI") +
  facet_grid(y_var ~ scenario, scales = "free_y") +
  labs(title = "Age Composition Change vs. Δχ² (Age)",
       x = expression(Delta ~ chi[Age]^2), y = "Δ (age group %)") +
  theme_lambert
```

### Figure — ΔGHI vs. Δχ²

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
chi_long <- analysis_demo %>%
  select(tract, scenario, change_green, delta_chi_age, delta_chi_race) %>%
  pivot_longer(cols = c(delta_chi_age, delta_chi_race),
               names_to = "chi_var", values_to = "chi_value") %>%
  mutate(chi_var = recode(chi_var,
    delta_chi_age = "Δχ² (Age)", delta_chi_race = "Δχ² (Race)"))

ggplot(chi_long, aes(x = chi_value, y = change_green)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.4, size = 0.7, color = "#2c7bb6") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6) +
  facet_grid(scenario ~ chi_var, scales = "free_x") +
  labs(title = "ΔGHI vs. Demographic Δχ²",
       x = "Δχ²", y = expression(Delta ~ "GHI")) +
  theme_lambert
```

---

## Reference

Lambert, M. R., Des Roches, S., Auerbach, D. A., Van Deynze, B., Behrens, S., Hale, R., & Pierce, K. B. (2025). Building the neighborhood for the trees: Illuminating win–wins for housing densification and nature. *Conservation Science and Practice*, 7(7), e70085.
